---
title: Genotyping checks
execute: 
  warning: false
  cache: true
---


```{r}
library(knitr)
library(data.table)
### library(scater)   ## scater:::.get_palette
library(DT)
library(ggpubr)
library(dplyr)

knitr::opts_chunk$set(cache=TRUE,echo=FALSE,message=FALSE,warning=FALSE,cache.lazy=FALSE)

## white background theme for all plots
theme_set(theme_bw())
```

There're 2 RNA files in the code from Frank, I'll just check the DNA and RNA file for now.

```{r}
samples_meta_file <- here::here("03_data/990_processed_data/001_snrnaseq/90_sample_info/samples_meta_merged_final.txt")

# GT_RNA_RNA_file <- "/Users/frankwessely/Desktop/Documents/Projects/BBB/Endo_10X_Main/Analysis/Genotyping/Comparisons/Compare_40_samples_RNA_RNA_raw/gtcheck_RNA_RNA_raw.txt"
 
# GT_RNA_RNA_filtered_file <- "/Users/frankwessely/Desktop/Documents/Projects/BBB/Endo_10X_Main/Analysis/Genotyping/Comparisons/Compare_40_samples_RNA_RNA_filtered/gtcheck_RNA_RNA_DP_20_800_qual_20.txt"
 
# GT_DNA_RNA_file <- "/Users/frankwessely/Desktop/Documents/Projects/BBB/Endo_10X_Main/Analysis/Genotyping/Comparisons/Compare_40_samples_DNA_RNA_filtered/gtcheck_40_samples_DP_20_800_qual_20.txt"
GT_DNA_RNA_file <- here::here("03_data/992_genotyping_data/04_compare_samples_dna_rna_filtered/gtcheck_40_samples_DP_20_800_qual_20.txt")
```

# Samples table

```{r}
#| label: samples_table

samples_meta <- fread(samples_meta_file, colClasses = c(donor = "character"))

samples_meta_print <- copy(samples_meta)
setnames(samples_meta_print, "Diagnosis", "______________________________Diagnosis______________________________")

samples_meta_print %>% DT::datatable(caption="Samples overview", escape=F, rownames = F, extensions = 'Buttons', options = list(dom = 'Blfrtip', buttons = list('colvis', list(extend = 'copy', exportOptions = list(columns=':visible')), list(extend = 'csv', exportOptions = list(columns=':visible')), list(extend = 'excel', exportOptions = list(columns=':visible')))))
```

# Genotyping: DNA vs RNA

* compare SNPs from 40 RNA BAM files to SNPs from donor Neurochip data
* percentage values indicate the percentage of SNPs matched based on gtcheck

***

```{r}
#| label: genotyping_DNA_RNA
#| fig-width: 9
#| fig-height: 7

GT <- fread(GT_DNA_RNA_file)
setnames(GT, c("S1", "S2", "n_1", "n_2", "n_different", "n_compared"))
GT <- GT[, pct_identity := 100 * (n_compared - n_different) / n_compared]

GT <- GT[, S1 := gsub("_var_raw", "", S1)]
GT <- GT[, S2 := gsub("_var_raw", "", S2)]

# samples_DNA <- c("1_NP002_2019","2_NP032_2018","3_NP070_2019","4_NP177_2013","5_NP003_2019","6_NP034_2014","7_NP076_2014","8_NP150_2014","9_NP008_2015","10_NP041_2019","11_NP077_2018","12_NP002_2019_R","13_NP011_2019","14_NP058_2019","15_NP080_2014","16_NP003_2019_R","17_NP016_2017","18_NP063_2019","19_NP128_2017","20_NP008_2015_R","21_NP018_2019","22_NP067_2019","23_NP142_2014","24_NP011_2019_R")
# 
# GT <- GT[, S1 := factor(S1, levels = samples_DNA)]
# GT <- GT[, S2 := factor(S2, levels = samples_order)]
# Extract parts of the sample labels
samples_order <- GT |>
  dplyr::select(S2) |>
  mutate(Prefix = substr(S2, 1, 1),
         NumericPart = as.integer(substr(S2, 3, nchar(S2)))) |>
  dplyr::distinct() |>
  dplyr::arrange(NumericPart) |>
  dplyr::pull(S2)

order_s1 <- GT %>%
  dplyr::group_by(S2) %>%
  dplyr::arrange(desc(pct_identity)) %>%
  dplyr::slice_head(n = 1) %>%
  dplyr::ungroup() |>
  dplyr::pull(S1)

GT <- GT |> 
  mutate(S2 = factor(S2, levels = samples_order))

# Reorder levels of S1 based on the calculated order
custom_order <- GT %>%
  mutate(S1 = factor(S1, levels = unique(order_s1)))
 
p <- ggplot(GT, aes(S1, S2))
p <- p + geom_tile(aes(fill = pct_identity))
p <- p + geom_text(aes(label = round(pct_identity)), colour="grey", size = 2.5)
p <- p + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0))
p <- p + xlab(NULL) + ylab(NULL)
p <- p + theme(legend.position="top")
p
```

Make ID dataframe to match neurchips to samples

```{r}
sample_ids <- data.frame(fraction_donor = samples_order, neurochip = order_s1)

# save sample IDs
readr::write_csv(sample_ids, here::here("03_data/990_processed_data/001_snrnaseq/90_sample_info/neurchip_sample_ids.csv"))
```


***

* same as above, reorder DNA donor samples on x-axis

***

```{r genotyping_DNA_RNA_2, fig.width=9, fig.height=7}
p <- ggplot(custom_order, aes(S1, S2))
p <- p + geom_tile(aes(fill = pct_identity))
p <- p + geom_text(aes(label = round(pct_identity)), colour="grey", size = 2.5)
p <- p + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0))
p <- p + xlab(NULL) + ylab(NULL)
p <- p + theme(legend.position="top")
p
```


***

```{r genotyping_DNA_RNA_table, cache=F}

GT %>% DT::datatable(caption="Genotyping, pairwise comparison, donor DNA (NeuroChip) vs RNA", escape=F, rownames = F, extensions = 'Buttons', options = list(dom = 'Blfrtip', buttons = list('colvis', list(extend = 'copy', exportOptions = list(columns=':visible')), list(extend = 'csv', exportOptions = list(columns=':visible')), list(extend = 'excel', exportOptions = list(columns=':visible')))))
```



***

# SessionInfo

```{r session_info, cache=F}

devtools::session_info()
```

