---
title: Perform Pseudobulk
execute:
  eval: true
#self-contained: true
code-fold: true
bibliography: references.bib
---

Note that I'll use the higher level annotations to start with.
This only involved aggregating the venous/capillary/arterial subtypes to a larger "Endothelial" category.

```{r}
here::i_am("04_data_analysis/001_snrnaseq_analysis/11_pseudobulk.qmd")
```

# Load packages

```{r}
source(here::here("04_data_analysis/990_code_libraries/02_library-calls-for-renv.R"))
```

```{r}
path <- here::here("04_data_analysis/990_code_libraries/pseudobulk")
#source necessary functions from other scripts
#source(here::here(path, "validate_input_parameters_de.R"))
source(here::here(path, "make_pseudobulk.R"))
source(here::here(path, "de_analysis.R"))
source(here::here(path, "plot_de_analysis.R"))
```

# Read in data

First we'll read in the scFlow output to a sce

```{r}
#| eval: true
# read data
#sce <- readr::read_rds(here::here("03_data/990_processed_data/001_snrnaseq/07_scflow_analysis/scflow-sce-annotated.rds"))
sce <- readr::read_rds(here::here("03_data/990_processed_data/001_snrnaseq/07_scflow_analysis/scflow-sce-annotated-highlevel.rds"))
#sce <- readr::read_rds(here::here("03_data/990_processed_data/001_snrnaseq/07_scflow_analysis/scflow-sce-updated_dim-35.rds"))
```

```{r}
# make count rownames match data rownames
rownames(sce@assays$RNA@counts) <- rownames(sce@assays$RNA@data)
sce$doner <- paste0("doner-", sce$individual)
```

# Average expression

```{r}
avg_exp <- AverageExpression(sce, group.by = c("ident", "doner"))
avg_exp <- avg_exp$RNA
 
avg_exp <- data.table(avg_exp, keep.rownames = TRUE)
setnames(avg_exp, "rn", "gene")

# pivot longer such that celltypes are separated out
avg_exp <- avg_exp %>%
  pivot_longer(
    cols = !gene,
    names_to = c("celltype", ".value"),
    names_pattern = "(.*)_(.*)"
  )
 
file_out <- here::here("03_data/990_processed_data/001_snrnaseq/08_pseudobulk/avg-expression-per-celltype-per-doner.csv")
#file_out <- here::here("03_data/990_processed_data/001_snrnaseq/08_pseudobulk/avg-expression-per-celltype-per-doner_35-dims.csv")
readr::write_csv(avg_exp, file_out)
 
## split AD and CT samples
# 
# avg_exp <- AverageExpression(sdata_endo, group.by = c("ident", "type"))
# avg_exp <- avg_exp$RNA
# 
# avg_exp <- data.table(avg_exp, keep.rownames = T)
# setnames(avg_exp, "rn", "gene")
# 
# file_out <- "/scratch/c.mpmfw/Endo_10X_main/Analysis/Expression_matrix/avg_expr_main_cell_type_40_samples_AD_vs_CT_v3.txt"
# write.table(avg_exp, file = file_out, sep = "\t", quote = F, row.names = F, col.names = T)
```

```{r}
# Make column with cluster and annotation combined
sce$cluster_annoated <- paste0("cluster_", sce$seurat_clusters, "_", Idents(sce))

avg_exp <- AverageExpression(sce, group.by = c("cluster_annoated", "doner"))
avg_exp <- avg_exp$RNA
 
avg_exp <- data.table(avg_exp, keep.rownames = TRUE)
setnames(avg_exp, "rn", "gene")

# pivot longer such that celltypes are separated out
avg_exp <- avg_exp %>%
  pivot_longer(
    cols = !gene,
    names_to = c("celltype", ".value"),
    names_pattern = "(.*)_(.*)"
  )
 
file_out <- here::here("03_data/990_processed_data/001_snrnaseq/08_pseudobulk/avg-expression-per-cluster-per-doner.csv")
readr::write_csv(avg_exp, file_out)
```

# Pseudobulk

Now we can subset to variables we want to aggregate across for the pseudobulk

```{r}
# aggregate read counts 
aggr_counts <- Seurat2PB(sce, sample="group", cluster="cluster_celltype")
dim(aggr_counts)
head(aggr_counts$samples, n=10L)
```

Convert back to single cell experiment

```{r}
sce <- Seurat::as.SingleCellExperiment(sce)
```

# Code from Alan

There's an interest in incorporating pseudobuilk into scflow itself, and a member of Nathan's lab (Alan) has written some code for performing pseudobulk on scflow outputs.

Here I'll play with his code and start thinking about how this could be incorporated into scflow

```{r}
#| eval: false
#| include: false

#Ran on R 4.3

#Get package dependencies

if (!require("BiocManager")) install.packages("BiocManager")
if (!require("RcppParallel")) install.packages("RcppParallel")
if (!require("pacman")) install.packages("pacman")
if (!require("qs")) install.packages("qs")
if (!require("SingleCellExperiment"))
  BiocManager::install("SingleCellExperiment")
if (!require("biomaRt")) BiocManager::install("biomaRt")
if (!require("ggplot2")) install.packages("ggplot2")
if (!require("cowplot")) install.packages("cowplot")
if (!require("viridis")) install.packages("viridis")
if (!require("ggrepel")) install.packages("ggrepel")
if (!require("Hmisc")) install.packages("Hmisc")
if (!require("edgeR")) BiocManager::install("edgeR")
if (!require("stats")) install.packages("stats")
if (!require("data.table")) install.packages("data.table")
if (!require("reshape2")) install.packages("reshape2")
if (!require("wesanderson")) install.packages("wesanderson")
if (!require("EnsDb.Hsapiens.v79")){

  #This package uses dependencies which are large so increase time for download
  options(timeout=2000)
  BiocManager::install("EnsDb.Hsapiens.v79")
}

#source the necessary function to run pb DE
source(here::here("04_data_analysis/990_code_libraries/pseudobulk/sc_cell_type_de.R"))

#library necessary packages
library(ggplot2)
library(data.table)
```

```{r}
#| eval: false
#| include: false

#load dataset and exclude endothelial cells to match Tsai group analysis

# Mathys_SCE <-
#   qs::qread("../data/sce.qs")

#remove endothelial cells to match Mathys 2019's approach
#Mathys_SCE <- Mathys_SCE[,!Mathys_SCE$allan_celltype=="Endo"]

# load data
sce <- readr::read_rds(here::here("03_data/990_processed_data/001_snrnaseq/07_scflow_analysis/scflow-sce.rds"))

#run DE analysis
sc_cell_type_de_return <-
  sc_cell_type_de(sce, design = ~ group,
                  pseudobulk_ID="manifest", celltype_ID="cluster_celltype",
                  folder = here::here("03_data/990_processed_data/001_snrnaseq/07_scflow_analysis/graphs/"),
                  verbose=TRUE)

#TO DO - add gene name to DEGs in results.......

#save results

#Cell Counts
data.table::fwrite(data.table(cell=names(
  sc_cell_type_de_return$celltype_counts),
  count=sc_cell_type_de_return$celltype_counts),
  "../results/cell_counts.csv")

#DEGs
data.table::fwrite(
  data.table::rbindlist(sc_cell_type_de_return$celltype_DEGs,idcol='Cell'),
  "../results/DEGs.csv")

#All Genes
data.table::fwrite(
  data.table::rbindlist(sc_cell_type_de_return$celltype_all_genes,idcol='Cell'),
  "../results/all_genes.csv")

#R object
save(sc_cell_type_de_return,file="../results/Mathys_pb_res.RData")
```

<!-- ## A more manual approach -->

<!-- Try extracting just the parts of `sc_cell_type_de.R` -->

```{r}
#| eval: false
#| include: false
# inputs
SCE <- sce
celltype_ID <- "ident"
design <- ~ group
pseudobulk_ID = "manifest"
y = NULL
pb_columns <- c("group", "manifest")
verbose = TRUE
control = "Control_V_oxford_dunno"
pval_adjust_method = "BH"
adj_pval=0.05
coef = NULL
folder = here::here("05_figures/001_snrnaseq_analysis_figures/01_pseudobulk/control_v_oxford")
```

```{r}
#| eval: false
#| include: false
#first format formula
design_txt <- paste0(deparse(design,width.cutoff = 500),collapse=',')

#make design formula minus anything before ~
formula <- as.formula(gsub(".*~","~",design_txt))

#if y not specified take last value in design matrix
if(is.null(y))
  y <- design_txt[[length(design_txt)]]

#Check if continuous or categorical variable to be modeled
y_contin <- FALSE
if(is.numeric(SCE[[y]]))
    y_contin <- TRUE
```

<!-- Get cell type counts -->

```{r}
#| eval: false
#| include: false
#get counts of each cell type
#counts(SCE) <- as.matrix(counts(SCE))

counts_celltypes <- SCE[[celltype_ID]]
counts_celltypes <-as.vector(table(counts_celltypes))
names(counts_celltypes) <- names(table(SCE[[celltype_ID]]))

# Get pseudobulk values
celltypes <- unique(SCE[[celltype_ID]])
```

```{r}
#| include: false
# counts by sample population
# cluster_count_by_sample <- table(sce$cluster_celltype, sce$prep) %>%
#  as.data.frame() %>%
#   pivot_wider(names_from = Var2, values_from = Freq) %>%
#   dplyr::rename(cell_type = Var1)
# 
# gt::gt(cluster_count_by_sample)
# 
# # Save data
# readr::write_csv(cluster_count_by_sample, here::here("03_data/990_processed_data/001_snrnaseq/07_scflow_analysis/celltype_counts_by_cellpop.csv"))
```

```{r}
#| eval: false
#| include: false
pb_dat <-
    lapply(celltypes,function(x)
        make_pseudobulk(SCE[,SCE[[celltype_ID]]==x],
                        pseudobulk_ID=pseudobulk_ID,
                        pb_columns=pb_columns))

names(pb_dat) <- celltypes
```

<!-- One can set the comparison here via the `control` argument. -->

```{r}
#| eval: false
#| include: false
#run edgeR LRT DE analysis
celltype_de <-
    de_analysis(pb_dat,formula,y_name=y,y_contin,coef,control,
                    pval_adjust_method, adj_pval, verbose)
```

```{r}
#| eval: false
#| include: false
#get sig DEGs for each
celltype_DEGs <- lapply(celltype_de, function(x) x[x$adj_pval<adj_pval,])
unique_genes <-lapply(celltype_DEGs, function(x) x$name)
unique_degs <- unique(unlist(unique_genes))

#if no DEGs found break but return the DE analysis scores still
if(length(unique_degs)==0)
    return(list("celltype_DEGs"=celltype_DEGs,
                "celltype_all_genes"=celltype_de,
                "celltype_counts"=counts_celltypes))

if(isTRUE(verbose))
    message(length(unique_degs)," unique DEGs foundacross all cell types")

celltype_all_genes_dt <- data.table::rbindlist(celltype_de,idcol = T)
setnames(celltype_all_genes_dt,".id","celltype")

celltype_DEGs_dt <- data.table::rbindlist(celltype_DEGs,idcol = T)
setnames(celltype_DEGs_dt,".id","celltype")
```

<!-- Changed palette arg in code to make this work -->

```{r}
#| eval: false
#| include: false
    if(!isFALSE(folder)){
        if(isTRUE(verbose))
            message("Plotting the results of differential expression analysis")

        #make plots for DE analysis
        plot_de_analysis(pb_dat,y,celltype_DEGs_dt,celltype_all_genes_dt,
                         counts_celltypes,folder)
    }
```

```{r}
#| eval: false
#| include: false
#get gene names for DEGs
genes <- unique(data.table::rbindlist(celltype_DEGs)$name)

gene_IDs <- ensembldb::select(EnsDb.Hsapiens.v79, keys= genes,
                              keytype = "GENEID",
                              columns = c("GENEID","SYMBOL"))

colnames(gene_IDs) <- c("ensembl_gene_id","hgnc_symbol")

gene_IDs <- data.table::as.data.table(gene_IDs)

data.table::setnames(gene_IDs,"ensembl_gene_id","name")

#remove any dups in the reference set - two names for one ENSEMBL ID
gene_IDs <- unique(gene_IDs,by="name")

celltype_DEGs <-
  lapply(celltype_DEGs,
         function(i){setDT(i);setkey(i,name);
           i[,gene_name:=gene_IDs[i, on=.(name),x.hgnc_symbol]];
           setnames(i,"name","ensembl_id");setnames(i,"gene_name","HGNC")})

#get gene names for all genes
genes <- unique(data.table::rbindlist(celltype_de)$name)

gene_IDs <- ensembldb::select(EnsDb.Hsapiens.v79, keys= genes,
                              keytype = "GENEID",
                              columns = c("GENEID","SYMBOL"))

colnames(gene_IDs) <- c("ensembl_gene_id","hgnc_symbol")

gene_IDs <- data.table::as.data.table(gene_IDs)

data.table::setnames(gene_IDs,"ensembl_gene_id","name")

#remove any dups in the reference set - two names for one ENSEMBL ID
gene_IDs <- unique(gene_IDs,by="name")

celltype_de <-
  lapply(celltype_de,
         function(i){setDT(i);setkey(i,name);
           i[,gene_name:=gene_IDs[i, on=.(name),x.hgnc_symbol]];
           setnames(i,"name","ensembl_id");setnames(i,"gene_name","HGNC")})

pseudobulk_results <- (list("celltype_DEGs"=celltype_DEGs,
            "celltype_all_genes"=celltype_de,
            "celltype_counts"=counts_celltypes))
```

<!-- Save pseudobulk results -->

```{r}
#| eval: false
#| include: false
readr::write_rds(pseudobulk_results, here::here("03_data/990_processed_data/001_snrnaseq/07_scflow_analysis/pseudobulk_results.rds"))
```

<!-- # try 2 -->

<!-- Try using `region` arg in `make_pseudobulk` -->


```{r}
#| eval: false
#| include: false

# inputs

SCE <- sce

celltype_ID <- "ident"

design <- ~ diagnosis

pseudobulk_ID = "manifest"

y = "diagnosis"

region = "prep"

pb_columns <- c("prep", "diagnosis", "manifest")

verbose = TRUE

control = "Control"

pval_adjust_method = "BH"

adj_pval=0.05

coef = "Case"

folder = here::here("03_data/990_processed_data/001_snrnaseq/07_scflow_analysis/graphs/")

```

```{r}
#| eval: false
#| include: false

#first format formula

design_txt <- paste0(deparse(design,width.cutoff = 500),collapse=',')

#make design formula minus anything before ~

formula <- as.formula(gsub(".*~","~",design_txt))

#if y not specified take last value in design matrix

if(is.null(y))

  y <- design_txt[[length(design_txt)]]

#Check if continuous or categorical variable to be modeled

y_contin <- FALSE

if(is.numeric(SCE[[y]]))

    y_contin <- TRUE

```

```{r}
#| eval: false
#| include: false
#get counts of each cell type

#counts(SCE) <- as.matrix(counts(SCE))

counts_celltypes <- SCE[[celltype_ID]]

counts_celltypes <-as.vector(table(counts_celltypes))

names(counts_celltypes) <- names(table(SCE[[celltype_ID]]))

# Get pseudobulk values

celltypes <- unique(SCE[[celltype_ID]])

pb_dat <-

    lapply(celltypes,function(x)

        make_pseudobulk(SCE[,SCE[[celltype_ID]]==x],

                        pseudobulk_ID=pseudobulk_ID,

                        pb_columns=pb_columns))

names(pb_dat) <- celltypes
```

```{r}
#| eval: false
#| include: false


#run edgeR LRT DE analysis

celltype_de <-

    de_analysis(pb_dat,formula,y_name=y,y_contin,coef,control,

                    pval_adjust_method, adj_pval, verbose)

#get sig DEGs for each

celltype_DEGs <- lapply(celltype_de, function(x) x[x$adj_pval<adj_pval,])

unique_genes <-lapply(celltype_DEGs, function(x) x$name)

unique_degs <- unique(unlist(unique_genes))

#if no DEGs found break but return the DE analysis scores still

if(length(unique_degs)==0)

    return(list("celltype_DEGs"=celltype_DEGs,

                "celltype_all_genes"=celltype_de,

                "celltype_counts"=counts_celltypes))

if(isTRUE(verbose))

    message(length(unique_degs)," unique DEGs foundacross all cell types")

celltype_all_genes_dt <- data.table::rbindlist(celltype_de,idcol = T)

setnames(celltype_all_genes_dt,".id","celltype")

celltype_DEGs_dt <- data.table::rbindlist(celltype_DEGs,idcol = T)

setnames(celltype_DEGs_dt,".id","celltype")

```

<!-- Changed palette arg in code to make this work -->

```{r}
#| eval: false
#| include: false

    if(!isFALSE(folder)){

        if(isTRUE(verbose))

            message("Plotting the results of differential expression analysis")

        #make plots for DE analysis

        plot_de_analysis(pb_dat,y,celltype_DEGs_dt,celltype_all_genes_dt,

                         counts_celltypes,folder)

    }

```

```{r}
#| eval: false
#| include: false

#get gene names for DEGs

genes <- unique(data.table::rbindlist(celltype_DEGs)$name)

gene_IDs <- ensembldb::select(EnsDb.Hsapiens.v79, keys= genes,

                              keytype = "GENEID",

                              columns = c("GENEID","SYMBOL"))

colnames(gene_IDs) <- c("ensembl_gene_id","hgnc_symbol")

gene_IDs <- data.table::as.data.table(gene_IDs)

data.table::setnames(gene_IDs,"ensembl_gene_id","name")

#remove any dups in the reference set - two names for one ENSEMBL ID

gene_IDs <- unique(gene_IDs,by="name")

celltype_DEGs <-

  lapply(celltype_DEGs,

         function(i){setDT(i);setkey(i,name);

           i[,gene_name:=gene_IDs[i, on=.(name),x.hgnc_symbol]];

           setnames(i,"name","ensembl_id");setnames(i,"gene_name","HGNC")})

#get gene names for all genes

genes <- unique(data.table::rbindlist(celltype_de)$name)

gene_IDs <- ensembldb::select(EnsDb.Hsapiens.v79, keys= genes,

                              keytype = "GENEID",

                              columns = c("GENEID","SYMBOL"))

colnames(gene_IDs) <- c("ensembl_gene_id","hgnc_symbol")

gene_IDs <- data.table::as.data.table(gene_IDs)

data.table::setnames(gene_IDs,"ensembl_gene_id","name")

#remove any dups in the reference set - two names for one ENSEMBL ID

gene_IDs <- unique(gene_IDs,by="name")

celltype_de <-

  lapply(celltype_de,

         function(i){setDT(i);setkey(i,name);

           i[,gene_name:=gene_IDs[i, on=.(name),x.hgnc_symbol]];

           setnames(i,"name","ensembl_id");setnames(i,"gene_name","HGNC")})

pseudobulk_results <- (list("celltype_DEGs"=celltype_DEGs,

            "celltype_all_genes"=celltype_de,

            "celltype_counts"=counts_celltypes))

```

<!-- Save pseudobulk results -->

```{r}
#| eval: false
#| include: false
readr::write_rds(pseudobulk_results, here::here("03_data/990_processed_data/001_snrnaseq/07_scflow_analysis/pseudobulk_results.rds"))
```

# Write from scratch

Here I'll try writing my own version of `de_analysis()`

I'll base it off the `edgeR` doc [here](https://www.bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf)

```{r}
sce$donor <- paste0("donor-", sce$individual)
levels(sce$diagnosis) <- c("AD", "Control")
```

```{r}
# inputs
SCE <- sce
celltype_ID <- "ident"
design <- ~ diagnosis
pseudobulk_ID = "manifest"
y = "diagnosis"
region = "prep"
#pb_columns <- c("prep", "diagnosis", "manifest")
pb_columns <- c("group", "manifest")
verbose = TRUE
control = "Control"
pval_adjust_method = "BH"
adj_pval=0.05
coef = "Case"
folder = here::here("03_data/990_processed_data/001_snrnaseq/07_scflow_analysis/graphs/")
```

```{r}
#first format formula
design_txt <- paste0(deparse(design,width.cutoff = 500),collapse=',')

#make design formula minus anything before ~
formula <- as.formula(gsub(".*~","~",design_txt))

#if y not specified take last value in design matrix
if(is.null(y))
  y <- design_txt[[length(design_txt)]]

#Check if continuous or categorical variable to be modeled
y_contin <- FALSE
if(is.numeric(SCE[[y]]))
    y_contin <- TRUE
```

Note that the pseudobulk compute takes a few minutes to run, so I'll disable eval on compilation.

```{r}
#| eval: false
#get counts of each cell type
#counts(SCE) <- as.matrix(counts(SCE))
counts_celltypes <- SCE[[celltype_ID]]
counts_celltypes <-as.vector(table(counts_celltypes))
names(counts_celltypes) <- names(table(SCE[[celltype_ID]]))

# Get pseudobulk values
celltypes <- unique(SCE[[celltype_ID]])
pb_dat <-
    lapply(celltypes,function(x)
        make_pseudobulk(SCE[,SCE[[celltype_ID]]==x],
                        pseudobulk_ID=pseudobulk_ID,
                        pb_columns=pb_columns))

names(pb_dat) <- celltypes

# the pseudobulk values take a while to compute, so I'll save the output
readr::write_rds(pb_dat, here::here("03_data/990_processed_data/001_snrnaseq/08_pseudobulk/pseudobulk_values.rds"))
```

```{r}
# load values
pb_dat <- readr::read_rds(here::here("03_data/990_processed_data/001_snrnaseq/08_pseudobulk/pseudobulk_values.rds"))
```

Hannah is after the pseudobulk values themselves, so I'll add the celltypes and save that data

```{r}
#| eval: false
pb_dat_per_individual <-
    lapply(celltypes,function(x)
        make_pseudobulk(SCE[,SCE[[celltype_ID]]==x],
                        pseudobulk_ID="individual"))

names(pb_dat_per_individual) <- celltypes

pseudo_vals <- map2(pb_dat_per_individual, names(pb_dat_per_individual), ~ .x$sumDat %>%
              as.data.frame() %>%
              dplyr::mutate(celltype = .y) %>%
              rownames_to_column("gene")) %>%
  purrr::list_rbind()

readr::write_csv(pseudo_vals, here::here("03_data/990_processed_data/001_snrnaseq/08_pseudobulk/pseudobulk_values_per_individual.csv"))
```

First part is prep for LRT, this is done per cell type via `purrr`

```{r}
# Get DGE list per celltype
y <- purrr::map(pb_dat, ~ edgeR::DGEList(counts = .x$sumDat,
                    group = .x$annot_pb$group))

# Calculate normalization factors to scale the raw library sizes.
y <- purrr::map(y, ~ edgeR::calcNormFactors(.x, method = 'TMM'))

# Build design matrix
design <- purrr::map(pb_dat, ~ model.matrix(~ group, data = .x$annot_pb))

# Maximizes the negative binomial likelihood to give the estimate of the
# common,trended and tagwise dispersions across all tags.
y <- purrr::map2(y, design, ~ edgeR::estimateDisp(.x, .y))

# LRT
fit <- purrr::map2(y, design, ~ edgeR::glmFit(.x, design = .y))
```

Now there's a need to think about what comparisons to run.

For now I think I'll write a more complex approach to run all possible comparisons from a arbitrarily long list

First I'll make a contrast matrix of all combinations

```{r}
# Get number group levels
ncls <- nlevels(sce$group)

get_contrasts <- function(design_matrix, group_vector) {
  
  num_groups <- nlevels(group_vector)
  contr <- rbind(matrix(1 / (1 - num_groups), num_groups, num_groups),
                 matrix(0, ncol(design_matrix) - num_groups, num_groups))

  diag(contr) <- 1
  contr[1,] <- 0

  rownames(contr) <- colnames(design_matrix)
  colnames(contr) <- paste0("cluster", levels(group_vector))

  return(contr)
}

contr <- purrr::map(design, ~ get_contrasts(.x, sce$group))
```

Then apply the LRT model per celltype

```{r}
apply_lrt_to_groups <-
  function(glmfit_object,
           contrast_matrix,
           group_levels) {
    lrt <-
      purrr::map(1:nrow(contrast_matrix), ~ {

        # Apply LRT per comparison
        lrt <-
          edgeR::glmLRT(glmfit_object, contrast = contrast_matrix[, .x])

        # Set name of comparison
        lrt$comparison <-
          paste0("group", group_levels[.x], "_vs_others")
        lrt
      })

    # Set names of comparison to list
    lrt <-
      purrr::set_names(lrt, map_chr(group_levels, ~ paste0("group", .x, "_vs_others")))
    
    return(lrt)
  }

lrt <- purrr::map2(fit, contr, ~ apply_lrt_to_groups(.x, .y, levels(sce$group)))
```

```{r}
# Check results
topTags(lrt$Endothelial$groupAD_P_oxford_dunno_vs_others)

# Get number of DE genes
dt <- purrr::map(lrt, ~ lapply(lapply(.x, decideTestsDGE), summary) %>%
                   do.call("cbind", .))
dt$Endothelial
```

Clean and extract results from lrt, including getting adjusted pvals

```{r}
clean_lrt_results <-
  function(lrt_results,
           pval_adjust_method = "BH",
           verbose = TRUE) {
    # Get results table
    pvals <- lrt_results$table
    
    # Add adj p-values
    pvals$adj_pval <- stats::p.adjust(pvals$PValue,
                                      method = pval_adjust_method)
    
    # Make rownames a column
    pvals$name <- rownames(pvals)
    rownames(pvals) <- NULL
    
    # Print number of DEGs found if verbose
    if (verbose) {
      numDEGs <- nrow(pvals[pvals$adj_pval < adj_pval, ])
      message(numDEGs, " DEGs found")
    }
    
    return(pvals)
}

lrt_clean <- map(lrt, ~ map(.x, ~clean_lrt_results(.x)))
```

Get significant genes and generate plots

- NOTE: plotting function needs to be updated - not currently working

```{r}
#| eval: false
#| include: false
celltype_de = lrt_clean$Endothelial
```

```{r}
get_sig_genes_and_plots <-
  function(celltype_de,
           pval_adjustment = 0.05,
           celltype_counts,
           plot_folder = FALSE,
           verbose = TRUE) {

    # get sig DEGs for each
    celltype_DEGs <-
      lapply(celltype_de, function(x)
        x[x$adj_pval < pval_adjustment, ])

    unique_genes <- lapply(celltype_DEGs, function(x)
      x$name)

    unique_degs <- unique(unlist(unique_genes))

    # if no DEGs found break but return the DE analysis scores still
    if (length(unique_degs) == 0)
      return(
        list(
          "celltype_DEGs" = celltype_DEGs,
          "celltype_all_genes" = celltype_de,
          "celltype_counts" = counts_celltypes
        )
      )

    if (isTRUE(verbose))
      message(length(unique_degs), " unique DEGs foundacross all cell types")

    celltype_all_genes_dt <-
      data.table::rbindlist(celltype_de, idcol = T)

    setnames(celltype_all_genes_dt, ".id", "celltype")

    celltype_DEGs_dt <-
      data.table::rbindlist(celltype_DEGs, idcol = T)

    setnames(celltype_DEGs_dt, ".id", "celltype")

    if(!isFALSE(plot_folder)){
        if(isTRUE(verbose))
            message("Plotting the results of differential expression analysis")

        #make plots for DE analysis
        # plot_de_analysis(pb_dat,y,celltype_DEGs_dt,celltype_all_genes_dt,
        #                  counts_celltypes,plot_folder)
    }

    return(celltype_DEGs)

}

lrt_sig <- map(
  lrt_clean,
  ~ get_sig_genes_and_plots(
    .x,
    celltype_counts = counts_celltypes,
    plot_folder = here::here("temp")
  )
)
```

The results are in a list of lists.
The first level is the celltypes and the next is the comparisons.

```{r}
# the top level is the celltypes
names(lrt_sig)
# the next level is the comparisons
names(lrt_sig$Endothelial)
head(lrt_sig$Endothelial$groupAD_P_oxford_dunno_vs_others)

# save results
readr::write_rds(lrt_sig, here::here("03_data/990_processed_data/001_snrnaseq/08_pseudobulk/pseudobulk_sig-genes_ad-vs-control-and-prep.rds"))
```

## Plots

Let's have a look at the results with some plots.

First I'll merge the different contrasts into a single dataframe.

```{r}
# merge contrasts
lrt_merged <- map(lrt_sig,
                  # add the contrast name as a column
                  ~ map2(
                    .x,
                    names(.x),
                    ~ .x %>%
                      dplyr::mutate(
                        contrast = .y,
                        # add direction of expression
                        deg_direction = ifelse(logFC > 0, "Up", "Down")
                      )
                  ) %>%
                    purrr::list_rbind())
```

```{r}
# set colour palette 
pal=c(wesanderson::wes_palette("Royal2"),
      wesanderson::wes_palette("Moonrise3"))
```

```{r}
#| eval: false
#| include: false

# this is supposed to use pb_dat I think
lrt_merged$Endothelial %>%
  dplyr::group_by(contrast) %>%
  # get top 3 genes by significance
  dplyr::slice_min(adj_pval, n = 3) %>%
  dplyr::ungroup() %>%
  dplyr::filter(name != "" & !is.na(deg_direction)) %>%
ggplot(aes(x = contrast, y = logFC,colour=deg_direction)) +
  geom_jitter(height=0) +
  stat_summary(fun.data = "mean_cl_normal",
               #aes(shape="mean"), 
               colour = "grey",
               geom = "crossbar",
               show.legend = T)+
  scale_shape_manual("", values=c("mean"="x"))+
  labs(y= "Sum pseudobulk expression counts (unnormalised)", 
          x = "Phenotype",
       colour="DEG direction") +
  facet_wrap(~ name, scales = "free_y")+
  theme_cowplot()+
  theme(axis.text = element_text(size=6),
       axis.text.x = element_text(angle = 45, hjust = 1, size = 6)) +
  #scale_colour_viridis(discrete = T)
  scale_colour_manual(values=pal)
```

```{r}
lrt_merged <- map2(lrt_merged, names(lrt_merged), ~ .x %>%
                     dplyr::mutate(celltype = .y)) %>%
  purrr::list_rbind()
```

### Counts of DEGs

```{r}
#| warning: false
lrt_merged %>%
  ggplot(aes(x = factor(deg_direction), fill = factor(celltype))) +
  geom_histogram(stat = "count") +
  facet_wrap(~factor(celltype)) +
  ggtitle("Cell type counts of DEGs")+
  labs(y= "Log2 Fold Change", x = "DEG direction", fill="Cell Type") +
  theme_cowplot()+
  theme(axis.text = element_text(size=9)) +
  scale_fill_viridis(discrete = T)
  #scale_fill_manual(values=pal)
```

```{r}
lrt_merged %>%
  ggplot(aes(x = contrast, y = logFC, fill = factor(contrast))) +
  geom_boxplot() +
  facet_wrap(~factor(celltype)) +
  ggtitle("Cell type Log2 Fold Change of DEGs") +#,
          # subtitle = paste0("Median LFC of Down and Up regulated DEGs: ",
          #                   round(degs_median_lfc$V1[[2]]*-1,2),", ",
          #                   round(degs_median_lfc$V1[[1]],2)))+
  labs(y= "Log2 Fold Change", x = "DEG direction", fill="Cell Type") +
  theme_cowplot()+
  theme(axis.text = element_text(size=6),
       axis.text.x = element_text(angle = 45, hjust = 1, size = 6)) +
  scale_fill_viridis(discrete = T, labels = c("AD P vs others", "AD V vs others", "Control P vs others", "Control V vs others"))
```

### Volcano plots

```{r}
# add colour identifier
cols <- c("Up" = "#FDE725FF", "Down" = "#440154FF")

lrt_merged %>%
    # scale FDR by log10
    ggplot(aes(x = logFC, y = -log10(adj_pval))) +
    geom_point(aes(colour = deg_direction), stat = "identity", size = .2) +
    facet_wrap(~ factor(celltype), scales = "free") +
    geom_hline(
        yintercept = -log10(0.05), colour = "#990000",
        linetype = "dashed", linewidth = .3
    ) +
    # geom_vline(xintercept = 0, colour="black", size=.3) +
    labs(y = "-log10(FDR)", x = "Log2 Fold Change", colour = "") +
    theme_cowplot() +
    theme(
        axis.text = element_text(size = 9), axis.text.x = element_blank(),
        legend.text = element_text(size = 10)
    ) +
    scale_colour_manual(values = cols) +
    geom_text_repel( # give top three genes per cell type by adjusted p value
        data = lrt_merged %>%
            dplyr::group_by(celltype) %>%
            dplyr::slice_min(adj_pval, n = 3),
        aes(label = name),
        size = 3,
        box.padding = unit(0.35, "lines"),
        point.padding = unit(0.3, "lines")
    )
```

# DESeq 2

```{r}
#| eval: true
# read data
#sce <- readr::read_rds(here::here("03_data/990_processed_data/001_snrnaseq/07_scflow_analysis/scflow-sce-annotated.rds"))
sce <- readr::read_rds(here::here("03_data/990_processed_data/001_snrnaseq/07_scflow_analysis/scflow-sce-annotated-highlevel.rds"))
```

```{r}
# make count rownames match data rownames
rownames(sce@assays$RNA@counts) <- rownames(sce@assays$RNA@data)
sce$doner <- paste0("doner-", sce$individual)
```

Let's first try doing pseudobulk by disease status

```{r}
# setup columns to aggregate by
sce$celltype <- Idents(sce)
# remove the "_" in the celltype names and replace with "-"
sce$celltype <- gsub("_", "-", sce$celltype)
sce$donor <- paste0("donor-", sce$individual)
levels(sce$diagnosis) <- c("AD", "Control")

# counts matrix individual level
cts <- AggregateExpression(sce, group.by = c("celltype", "donor", "diagnosis"),
                    assays = "RNA",
                    slot = "counts",
                    return.seurat = FALSE)
cts <- cts$RNA
cts[1:5,1:10]
```

Now we need to do some data manipulation to get the count and column data in the format DESeq2 requires

```{r}
# transpose
cts_t <- t(cts) %>%
  as.data.frame()

split_rows <- gsub('_.*', '', rownames(cts_t))

cts_split <- split.data.frame(cts_t,
                              f = factor(split_rows))

cts_split_m <- lapply(cts_split, function(x) {
  t(x)
})

# Function to modify column names
modify_column_names <- function(df, number_of_groups = 2) {
  df <- as.data.frame(df)
  # get the number of "_" seperations in the names
  number_of_splits <- length(strsplit(as.character(names(df)), "_")[[1]])
  # get the last n - 1 groups
  cols <-
    map((number_of_splits - (number_of_groups - 1)):number_of_splits,
        ~ sapply(strsplit(as.character(names(
          df
        )), "_"), "[", .x))
  names(df) <- do.call(paste, c(cols, sep = "_"))
  
  return(df)
}
cts_split_m <- map(cts_split_m, modify_column_names)
```

```{r}
cts_split_m$`Neuron-ex`[1:10,1:2]
```

```{r}
# get the col data
colData <- map(
  cts_split_m,
  ~ data.frame(
    samples = colnames(.x),
    condition = ifelse(grepl("Control", colnames(.x)), "Control", "AD")
  ) %>%
    column_to_rownames(var = "samples")
)
```

```{r}
# check that count and col data cols/rows match
map2_lgl(cts_split_m, colData, ~ ncol(.x) == nrow(.y))
```

```{r}
# get DESeq2 object
dds <- map2(cts_split_m, colData, ~ DESeqDataSetFromMatrix(countData = .x,
                                                           colData = .y,
                                                           design = ~ condition))
```

```{r}
# run DESeq2
dds <- map(dds, DESeq)
```

We can see the result names

```{r}
resultsNames(dds$Astro)
```

And then extract these results

```{r}
res <- map(dds, ~results(.x, name = "condition_Control_vs_AD"))
head(res$Astro)
```

```{r}
# merge celltype dataframes
res_merged <- map2(res, names(res), ~ as.data.frame(.x) %>%
                     rownames_to_column("gene") %>%
                     dplyr::mutate(celltype = .y, deg_direction = ifelse(log2FoldChange > 0, "Up", "Down"))) %>%
  purrr::list_rbind()
```

```{r}
# save results
readr::write_rds(res_merged, here::here("03_data/990_processed_data/001_snrnaseq/08_pseudobulk/pseudobulk_ad-vs-control_deseq2.rds"))
```

## Plots

```{r}
#| warning: false
res_merged %>%
  ggplot(aes(x = factor(deg_direction), fill = factor(celltype))) +
  geom_histogram(stat = "count") +
  facet_wrap(~factor(celltype)) +
  ggtitle("Cell type counts of DEGs - Control VS AD")+
  labs(y= "Log2 Fold Change", x = "DEG direction", fill="Cell Type") +
  theme_cowplot()+
  theme(axis.text = element_text(size=9)) +
  scale_fill_viridis(discrete = T)
  #scale_fill_manual(values=pal)
```

### Volcano plots

```{r}
# add colour identifier
cols <- c("Up" = "#FDE725FF", "Down" = "#440154FF")

res_merged %>%
  # filter to significant hits
  dplyr::filter(padj < 0.05) %>%
  # scale FDR by log10
  ggplot(aes(x = log2FoldChange, y = -log10(padj))) +
  geom_point(aes(colour = deg_direction), stat = "identity", size = .2) +
  facet_wrap( ~ factor(celltype), scales = "free") +
  geom_hline(
    yintercept = -log10(0.05),
    colour = "#990000",
    linetype = "dashed",
    linewidth = .3
  ) +
  # geom_vline(xintercept = 0, colour="black", size=.3) +
  labs(y = "-log10(FDR)", x = "Log2 Fold Change", colour = "") +
  theme_cowplot() +
  theme(
    axis.text = element_text(size = 9),
    axis.text.x = element_blank(),
    legend.text = element_text(size = 10)
  ) +
  scale_colour_manual(values = cols) +
  geom_text_repel(
    # give top three genes per cell type by adjusted p value
    data = res_merged %>%
      dplyr::group_by(celltype) %>%
      dplyr::slice_min(padj, n = 3),
    aes(label = gene),
    size = 3,
    box.padding = unit(0.35, "lines"),
    point.padding = unit(0.3, "lines")
  )
```


```{r}
#| eval: false
#| include: false
        #slight diff approach for contin or cat
        if(y_contin){#need to specify coefficient target as not done in design
            test <- edgeR::glmLRT(fit,coef=y_name)
        }
        else{#categorical i.e. case/control, no need
            if(!is.null(coef)){#if user inputted value for case samples
                test <- edgeR::glmLRT(fit,coef=length(new_levels))
            }
            else{#they didn't direction may not be as they expect
                message(paste0("WARNING: No coefficient passed so direction ma",
                               "y not relate to desired variable"))
                test <- edgeR::glmLRT(fit)
            }
        }

        pvals <- test$table

        #add adj p-values

        pvals$adj_pval <- stats::p.adjust(pvals$PValue,

                                          method = pval_adjust_method)

        pvals$name <- rownames(pvals)

        rownames(pvals) <- NULL

        if(verbose){

            numDEGs <- nrow(pvals[pvals$adj_pval<adj_pval,])

            message(numDEGs," DEGs found")

        }

        DE_ct_rtrn[[ct_i]] <- pvals

```
